Final paper: Towards Proving Termination of \texttt{zelda-mosaic} Algorithms under Conditions on the Input

### 2021-05-02 paper/paper.tex
paper: proof-reading pass

@@

\section{Problem Description}
Many non-trivial programs make use of arrays, and our \texttt{zelda-mosaic} program is no different. Because our goal was to prove the termination of the program when given a well-formed input and the core data structure modified throughout the program was an image (represented as an array), it is helpful to first describe the array operations used and the invariants of an array that contribute to ``well-formed-ness".

\subsection{Array Usage}
\matlab's support for multi-dimensional arrays makes it amenable to image-processing tasks such as ours. The algorithms we wrote are described at high-level in pseudocode in Figures \figurename~\ref{alg:v1}, \ref{alg:v2}, and \ref{alg:v3}, but here we explain them in layman's terms to help motivate some features of our proof.\\

As described in our pseudocode, the inputs to our program were a piece of \textit{key art} to be recomposed with \textit{thumbnails} of a certain \textit{size}, defined as the width and height of the thumbnails (all have been coerced to be square-shaped). The original representation of \textit{key art} in \matlab is a two-dimensional array (with indices corresponding to a particular pixel's row and column) containing an additional array of three values - one each for the red, green, and blue color channels. So far, this is an array of three dimensions. We then divide this image into \textit{chunks}, where each \textit{chunk} will be replaced by a \textit{thumbnail} in the final product. For \textit{key art} of dimensions $m \times n$ (with $m$ corresponding to and $n$ corresponding to columns), each image will be divided into $(m / size) \times (n / size)$ \textit{chunks}, which will each be replaced by a \textit{thumbnail} as appropriate. These \textit{chunks}, each of size $size \times size$, are then stored in a two-dimensional array, with each \textit{chunk} retaining its original placement in the \textit{key art}. As a result we now have a five-dimensional array, and in order to access a particular value from a color channel, we need the following values:

\begin{itemize}
\item the \textit{chunk row} it is in
\item the \textit{chunk column} it is in
\item the \textit{x-offset} of the pixel into its chunk
\item the \textit{y-offset} of the pixel into its chunk
\item the \textit{color channel} desired out of red, green, and blue
\end{itemize}

Thus, once a piece of \textit{key art} is successfully divided into \textit{chunks}, the image is effectively represented as an array whose dimensions correspond to the above values. The \textit{chunks} are iterated through in row-major order, gradually being replaced in-place by \textit{thumbnails}. Consequently, the state of our core array maintains some matrix-like properties that are supported by \matlab\@ arrays, but not particular to \matlab\@ in general:

\begin{itemize}
    \item The shape of a matrix \textbf{$M$} can be described by a list of \textit{d} dimensions, where each dimension's value $v_i$ refers to the number of elements held along that dimension. For example, an array with dimensions $3 \times 4 \times 2$ contains 3 arrays, each of which contains 4 arrays of 2 elements. $v_1 = 3$, $v_2 = 4$, and $v_3 = 2$.
    \item Once the shape of \textbf{$M$} is defined and all of the elements have been populated, the shape will not change and all of the elements will remain populated. For example, the aforementioned $3 \times 4 \times 2$ array must always contain 3 arrays, each of which contains 4 arrays of 2 elements.
    \item Each $i$-th dimension from $1$ to $d-1$ must contain $v_i$ elements, each of which is an array of length $v_{i+1}$; in other words, the length of each element in a particular dimension must be identical to one another.
\end{itemize}

In summary, these properties enforce a notion of ``well-formed-ness" that ensures that the size $v_i$ of each dimension remains fixed and that all element in a particular dimension $i$ have the same length as one another. This property is generally true of matrices, hence the references to both an ``arrays" and ``matrices" in this section. Unless otherwise noted, ``array" refers to the generic \matlab\@ object and ``matrix" refers to an ``array" with these properties when discussed in the scope of this paper.

\subsection{Operations} \matlab\@ supports a number of operations that can be used to manipulate arrays, but only a subset are used in our program. To illustrate the operations used, we begin with the following visualization of a three-dimensional array \textbf{$E$} with dimensions $3 \times 4 \times 2$, as it is used in \matlab:\\

@@

\begin{figure}[h!]
    \centering
    \begin{tikzpicture}[every node/.style={anchor=north east,fill=white, minimum width=.7cm}]
    \matrix (mA) [draw,matrix of math nodes]
    {
    1 & 3 & 5 & 7 \\
    2 & 4 & 6 & 8 \\
    };

@@

    \draw[dashed](mA.north east)--(mC.north east);
    \draw[dashed](mA.north west)--(mC.north west);
    \draw[dashed](mA.south east)--(mC.south east);
    \end{tikzpicture}

\end{figure}

\subsubsection{Element Access}
For the matrix \textbf{$E$}, each element would be indexed according to the following mapping. Under this scheme, for example, the element $8$ in \textbf{$E$} would have the index $(1, 4, 2)$. This indexing scheme is the general pattern used to access one particular element out of a multi-dimensional array. Note that arrays in \matlab\@ have indices starting at 1.

\begin{figure}[h!]
    \centering
    \begin{tikzpicture}
    \def\xs{1} %shift in x direction
    \def\ys{1.5} %shift in y direction
    \def\nm{3} % number of 2d matrices in the 3d matrix
    \foreach \x in {1,2,...,\nm}
    {
    \matrix [draw, % for the rectangle border
             fill=white, % so that it is not transparent
             ampersand replacement=\&] %see explanation
    (mm\x)%give the matrix a name
    at(-\x * \xs, -\x * \ys) %shift the matrix
    {
        \node {(\x,1,1)}; \& \node {(\x,2,1)}; \& \node {(\x,3,1)}; \& \node {(\x,4,1)};\\
        \node {(\x,1,2)}; \& \node {(\x,2,2)}; \& \node {(\x,3,2)}; \& \node {(\x,4,2)};\\
    };
    }

    \draw [dashed,gray](mm1.north west) -- (mm\nm.north west);
    \draw [dashed,gray](mm1.north east) -- (mm\nm.north east);
    \draw [dashed,gray](mm1.south east) -- (mm\nm.south east);
    \end{tikzpicture}
\end{figure}

\subsubsection{Linearized Indexing} \label{linearizedindexing}
Note that in $E$, elements are ordered in ascending order with respect to their indexing. First, sub-array $(1)$ is populated, and $(1, 1)$ is the first array within it, so it is filled with elements $1$ and $2$. Array $(1, 2)$ is then populated by elements $3$ and $4$. After all four of \textbf{$E$}'s sub-arrays have been populated, sub-array $(2)$ is populated, and so on until the array is full.

Given the numbering scheme in $E$, one can imagine a ``flattened" version of $E$ that is simply an ordered one-dimensional array containing all of $E$'s elements; simply put, an array containing the elements $1$ to $24$, in numerical order. \matlab\@ actually allows arrays of arbitrarily many dimensions to be logically ``flattened" during accesses so that elements can be accessed with one index, like in a one-dimensional array. We found this property to be somewhat interesting and challenging to model, and we refer to the ``flattening" of a matrix as ``linearization".

The process of linearizing a matrix is perhaps
better explained with a two-dimensional example. Imagine a general two-dimensional $3 \times 4$ matrix, such as the one below, where:
\begin{equation*}
    \begin{bmatrix}
        a & b & c & d \\
        e & f & g & h \\
        i & j & k & l
    \end{bmatrix}
\end{equation*}
To linearize the matrix, we label each value with an increasing index from the
top-left, left-to-right, top-to-bottom. For example:
\begin{equation*}
    \begin{bmatrix}
        a_1 & b_2 & c_3 & d_4 \\
        e_5 & f_6 & g_7 & h_8 \\
        i_9 & j_{10} & k_{11} & l_{12}
    \end{bmatrix}
\end{equation*}
Finally, we create a vector, a one-dimensional matrix, by reading the elements
off in order:
\setcounter{MaxMatrixCols}{20}
\begin{equation*}
    \begin{bmatrix}
        a & b & c & d & e & f & g & h & i & j & k & l
    \end{bmatrix}
\end{equation*}
This process generalizes to higher dimensions. The length of the resulting
vector is always the product of the sizes of the dimensions. To see this, note
that this product counts the number of elements in the matrix and that the
linearized matrix has the same number of elements in a single dimension.

Under this scheme, any element in a matrix can be accessed via its position in the linearized version of that matrix. For example, the element $12$ in \textbf{$E$} can be accessed by the index $(12)$ (note that 12 refers to the index, not the elements) and in the previous example, the element $f$ can be accessed by the index $(6)$.

\subsubsection{Ranged Indexing} \label{rangedindexing}
Broadly speaking, \matlab\@ supports the notion of a range of numbers specified by a starting number and a stopping number, which includes every number between the starting number and the stopping number (inclusive). This abstraction can be used to access or assign elements within a sub-matrix of a particular matrix. For example, suppose in matrix $E$ you only cared about the first matrix in the first dimension, the second/third/fourth columns within the second dimension, and the second row within the third dimension. This sub-matrix can be specified by the index $(1, 2:4, 2)$ and would return the matrix:
\begin{equation*}
    \begin{bmatrix}
        4 & 6 & 8
    \end{bmatrix}
\end{equation*}

Suppose that you wanted to get the second/third/fourth columns within the second dimension and second row within the third dimension for \textit{all three} of the outer-most dimension of \textbf{$E$} but did not know a priori how many elements it had. \matlab\@ supports an ``inclusive range" operator, represented by a colon \textbf{:}, which is taken to mean the range encompassing all values in a particular dimension. The sub-matrix described here would be specified by the index $(:, 2:4, 2)$ and represents the following matrix:

\begin{figure}[h!]
    \centering
    \begin{tikzpicture}[every node/.style={anchor=north east,fill=white, minimum width=.7cm}]
    \matrix (mA) [draw,matrix of math nodes]
    {
    4 & 6 & 8 \\
    };

@@

    \draw[dashed](mA.north east)--(mC.north east);
    \draw[dashed](mA.north west)--(mC.north west);
    \draw[dashed](mA.south east)--(mC.south east);
    \end{tikzpicture}

\end{figure}

\subsubsection{Matrix Operations} \label{matrixoperations}
Details about \matlab\@ arrays pertaining to indexing are perhaps some of their most interesting features, but we also use some typical matrix operators in our program. Generally speaking, matrix operators can be generalized to any number of dimensions, as long as the dimensions of the arrays agree for the operation in question. There are two broad classes of matrix operators that we make use of:

\begin{itemize}
    \item \textbf{Element-wise}: For two matrices of the same dimensionality, the operator is applied on each of the two elements containing the same index in the two matrices.
    \item \textbf{All Elements}: For an array and a scalar, the operation and scalar (e.g. add 1) is applied to every element in the matrix.
\end{itemize}

\subsection{Modeling the Problem}
Aside from the core matrix operations, we anticipated that we would need to model the state of our program at all points of execution since our goal was to prove termination. To this end, we would need to identify all of the syntax that we use in our program and specify their associated semantics. The upshot of modeling our program in this way would be the ability to remains to prove its termination of our given that certain preconditions hold.

@@

When we started this project, we had in mind that the preconditions would be statements about the program's inputs; for example, ``the length and width of \textit{key art} are always multiples of the \textit{thumbnails'} size". Ultimately, we did not get far enough in modeling our program to devise a set of pre-conditions, but the point remains that we needed to be able to model the control flow of our entire program in order to do this.

\section{Design / Approach}

@@

Much of the abstract syntax is relatively self-explanatory, so complete
definitions are elided here. However, we wish to draw attention to some of the
less apparent and more interesting syntax present in \mmatlab\@:

@@

    \item \textsf{AddExp}: While this is nominally a regular addition operator,
        \matlab's addition is interesting in that it has overloaded parameter
        types and expected behavior for each. The addition of two numbers is a
        number and the addition of two matrices results in element-wise
        addition, as one might expect. However, the sum of a number and a
        matrix, is a matrix with all elements incremented by the number. Similar
        behaviors hold for \textsf{MultExp} and \textsf{DivExp}. See Section \ref{matrixoperations} for the distinction between these operators.

    \item \textsf{RangeExp}: Like Python, \matlab\ has an expression where a
        starting number and stopping number may be provided, generating an
        inclusive list of all numbers between them. This is discussed in Section \ref{rangedindexing}

    \item \textsf{IndexExp}: In support of the RangeExp expression, the syntax to access elements from an array is overloaded in \matlab\@. They

@@

As mentioned in \S~\ref{S:design_dependent}, our original formulation of
matrices as nested dependent \texttt{Vector.t}s quickly grew in complexity.
We abandoned that approach in favor of a type that is not guaranteed to behave
well, but whose definition makes defining operations more
natural~\cite{SO_2021_2}. This approach removes the proofs that dependent types
provide automatically, so we have to give those ourselves.

@@

\paragraph{Linearizing a matrix} See \S~\ref{S:design_linearize} for more
details on matrix linearization. To define linearization in Coq, we use a
two-stage unpacking as with other matrix operations. The function
\(\func{linearize'}\) turns scalars into a singleton list and recursively
flat-maps itself over matrices to produce the one-dimensional matrix as a list.
The function \(\func{linearize}\) unpacks a matrix, then transforms the contents
by wrapping the result of mapping \(\func{linearize'}\) into \(\func{Scalar}\)s
in a \(\func{Matrix}\) and transforms the shape by taking the product of the
(normalized) shape of the matrix. Recall that the length of the linearized
vector is the product of the sizes of all the dimensions. The definitions are
spelled out below:

@@

\paragraph{Linearly indexing a matrix} Given the operation to linearise a
matrix, one-based linearly indexing is easy. See \S~\ref{S:design_linear_index}
for details on linear indexing. The function \(\func{nth}\) takes a matrix \(m\)
and an index \(n\); it returns the one-based linearly indexed content of \(m\)
at \(n\) (wrapped in the option constructor \(\func{Some}\)) if \(n\) is in
bounds relative to \(m\). When \(n\) is out of bounds, the result is
\(\func{None}\). For one-based indexing with functions that naturally use
zero-based indexing, we use \(n-1\). We should note that the current version of
\(\func{nth}\) has the subtle bug that indexing at \(n=0\) currently returns the
same as indexing at \(n=1\), when it should in fact be out of bounds and thus
\(\func{None}\)---a simple \texttt{if} or \texttt{match} guard would cover this
case.

@@

\begin{figure}[!t]
    \textbf{Input}: image \(key\_art\), images \(thumbnails\), integer \(size\) \\
    \textbf{Output}: image \(tiles\)
    \begin{algorithmic}
        \STATE{\(tiles \gets \textrm{MAT2TILES}(key\_art, size, size)\)}
        \FOR{\(tile \in tiles\)}
            \FOR{\(thumbnail, i \in thumbnails, [1..|thumbnails|]\)}
                \STATE{\(mses_i \gets \textrm{immse}(tile, thumbnail)\)}
            \ENDFOR
            \STATE{\(best\_indices \gets \textrm{find} (mses = \min mses)\)}
            \STATE{\(best\_index \gets best\_indices_0\)}
            \STATE{\(tile \gets thumbnails_{best\_index}\)}
        \ENDFOR
    \RETURN{\(tiles\)}
    \end{algorithmic}
    \caption{Algorithm: Version 1}\label{alg:v1}
\end{figure}

\begin{figure}[!t]
    \textbf{Input}: image \(key\_art\), images \(thumbnails\), integer \(size\) \\
    \textbf{Output}: image \(tiles\)
    \begin{algorithmic}
        \STATE{\(tiles \gets \textrm{MAT2TILES}(key\_art, size, size)\)}
        \STATE{\(used \gets \emptyset\)}
        \STATE{\(threshold \gets 1\)}
        \IF{\(|thumbnails| < |tiles|\)}
            \STATE{\(threshold \gets \lceil{|tiles| / |thumbnails|}\rceil\)}
        \ENDIF
        \FOR{\(tile \in tiles\)}
            \FOR{\(thumbnail, i \in thumbnails, [1..|thumbnails|]\)}
                \STATE{\(mses_i \gets \textrm{immse}(tile, thumbnail)\)}
            \ENDFOR
            \STATE{\(best\_indices \gets \textrm{find} (mses = \min mses)\)}
            \STATE{\(best\_index \gets best\_indices_0\)}
            \WHILE{\(used_{best\_index} \ge threshold\)}
                \STATE{\(mses_{best\_index} \gets \infty\)}
                \STATE{\(best\_indices \gets \textrm{find} (mses = \min mses)\)}
                \STATE{\(best\_index \gets best\_indices_0\)}
            \ENDWHILE
            \STATE{increment \(used_{best\_index}\)}
            \STATE{\(tile \gets thumbnails_{best\_index}\)}
        \ENDFOR
        \RETURN{\(tiles\)}
    \end{algorithmic}
    \caption{Algorithm: Version 2}\label{alg:v2}
\end{figure}

\begin{figure}[!t]
    \textbf{Input}: image \(key\_art\), images \(thumbnails\), integer \(size\) \\
    \textbf{Output}: image \(tiles\)
    \begin{algorithmic}
        \STATE{}\COMMENT{Repeat Version 2}
        \STATE{\(tiles \gets \textrm{Version2}(key\_art, thumbnails, size)\)}
        \STATE{\(tiles \gets \textrm{imgaussfilt}(tiles)\)}
        \STATE{\(strip\_size \gets \lfloor{size / 8}\rfloor\)}
        \STATE{}\COMMENT{Blur by averaging}
        \FOR{horizontal \(strip \in tiles\)}
            \STATE{\(strip \gets \textrm{mean}({strip})\)}
        \ENDFOR
        \FOR{vertical \(strip \in tiles\)}
            \STATE{\(strip \gets \textrm{mean}({strip})\)}
        \ENDFOR
        \RETURN{\(tiles\)}
    \end{algorithmic}
    \caption{Algorithm: Version 3}\label{alg:v3}
\end{figure}

>>>>

\begin{figure}[htp]
    \textbf{Input}: image \(key\_art\), images \(thumbnails\), integer \(size\) \\
    \textbf{Output}: image \(tiles\)
    \begin{algorithmic}
        \STATE{\(tiles \gets \textrm{MAT2TILES}(key\_art, size, size)\)}
        \FOR{\(tile \in tiles\)}
            \FOR{\(thumbnail, i \in thumbnails, [1..|thumbnails| ]\)}
                \STATE{\(mses_i \gets \textrm{immse}(tile, thumbnail)\)}
            \ENDFOR
            \STATE{\(best\_indices \gets \textrm{find} (mses = \min mses)\)}
            \STATE{\(best\_index \gets best\_indices_0\)}
            \STATE{\(tile \gets thumbnails_{best\_index}\)}
        \ENDFOR
    \RETURN{\(tiles\)}
    \end{algorithmic}
    \caption{Algorithm: Version 1}\label{alg:v1}
\end{figure}

\begin{figure}[htp]
    \textbf{Input}: image \(key\_art\), images \(thumbnails\), integer \(size\) \\
    \textbf{Output}: image \(tiles\)
    \begin{algorithmic}
        \STATE{\(tiles \gets \textrm{MAT2TILES}(key\_art, size, size)\)}
        \STATE{\(used \gets \emptyset\)}
        \STATE{\(threshold \gets 1\)}
        \IF{\(|thumbnails| < |tiles|\)}
            \STATE{\(threshold \gets \lceil{|tiles| / |thumbnails|}\rceil\)}
        \ENDIF
        \FOR{\(tile \in tiles\)}
            \FOR{\(thumbnail, i \in thumbnails, [1..|thumbnails| ]\)}
                \STATE{\(mses_i \gets \textrm{immse}(tile, thumbnail)\)}
            \ENDFOR
            \STATE{\(best\_indices \gets \textrm{find} (mses = \min mses)\)}
            \STATE{\(best\_index \gets best\_indices_0\)}
            \WHILE{\(used_{best\_index} \ge threshold\)}
                \STATE{\(mses_{best\_index} \gets \infty\)}
                \STATE{\(best\_indices \gets \textrm{find} (mses = \min mses)\)}
                \STATE{\(best\_index \gets best\_indices_0\)}
            \ENDWHILE
            \STATE{increment \(used_{best\_index}\)}
            \STATE{\(tile \gets thumbnails_{best\_index}\)}
        \ENDFOR
        \RETURN{\(tiles\)}
    \end{algorithmic}
    \caption{Algorithm: Version 2}\label{alg:v2}
\end{figure}

\begin{figure}[htp]
    \textbf{Input}: image \(key\_art\), images \(thumbnails\), integer \(size\) \\
    \textbf{Output}: image \(tiles\)
    \begin{algorithmic}
        \STATE{}\COMMENT{Repeat Version 2}
        \STATE{\(tiles \gets \textrm{Version2}(key\_art, thumbnails, size)\)}
        \STATE{\(tiles \gets \textrm{imgaussfilt}(tiles)\)}
        \STATE{\(strip\_size \gets \lfloor{size / 8}\rfloor\)}
        \STATE{}\COMMENT{Blur by averaging}
        \FOR{horizontal \(strip \in tiles\)}
            \STATE{\(strip \gets \textrm{mean}({strip})\)}
        \ENDFOR
        \FOR{vertical \(strip \in tiles\)}
            \STATE{\(strip \gets \textrm{mean}({strip})\)}
        \ENDFOR
        \RETURN{\(tiles\)}
    \end{algorithmic}
    \caption{Algorithm: Version 3}\label{alg:v3}
\end{figure}

@@

\section{Problem Description}

Many non-trivial programs make use of arrays, and our \texttt{zelda-mosaic}
program is no different. Because our goal was to prove the termination of the
program when given a well-formed input and the core data structure modified
throughout the program was an image (represented as an array), it is helpful to
first describe the array operations used and the invariants of an array that
contribute to the \emph{well-formed} property.

\subsection{Array Usage}

\matlab's support for multi-dimensional arrays makes it amenable to
image-processing tasks such as ours. The algorithms we wrote are described at
high-level in pseudocode in \figurename{s}~\ref{alg:v1},~\ref{alg:v2},
and~\ref{alg:v3}, but here we explain them in layman's terms to help motivate
some features of our proof.

As described in our pseudocode, the inputs to our program were a piece of
\emph{key art} to be recomposed with \emph{thumbnails} of a certain \emph{size},
which is the width and height of the square thumbnails. The original
representation of the key art in \matlab\ is a two-dimensional array (with
indices corresponding to a particular pixel's row and column) containing an
additional array of three values---one each for the red, green, and blue color
channels. In total, the array has three dimensions. We then divide this image
into \emph{chunks}, where each chunk will be replaced by a
thumbnail in the final product. For key art of dimensions $m
\times n$ (with $m$ corresponding to rows and $n$ corresponding to columns), each
image will be divided into $(m / \var{size}) \times (n / \var{size})$ chunks, which
will each be replaced by a thumbnail as appropriate. These chunks,
each of size $\var{size} \times \var{size}$, are then stored in a two-dimensional array,
with each chunk retaining its original placement in the key art.
As a result we now have a five-dimensional array, and in order to access a
particular value from a color channel, we need the following values:

\begin{itemize}
\item the \emph{chunk row} it is in
\item the \emph{chunk column} it is in
\item the \emph{x offset} of the pixel into its chunk
\item the \emph{y offset} of the pixel into its chunk
\item the \emph{color channel} desired out of red, green, and blue
\end{itemize}

Thus, once a piece of key art is successfully divided into chunks, the image is
effectively represented as an array whose dimensions correspond to the above
values. The chunks are iterated through in row-major order, gradually being
replaced in-place by thumbnails. Consequently, the state of our core array
maintains some matrix-like properties that are supported by \matlab\ arrays,
but not particular to \matlab\ in general:

\begin{itemize}
    \item The shape of a matrix \(M\) can be described by a list of \(d\)
        dimensions, where each dimension's value \(v_i\) refers to the number of
        elements held along that dimension. For example, an array with
        dimensions \(3 \times 4 \times 2\) contains 3 arrays, each of which
        contains 4 arrays of 2 elements. The dimensions list contains \(v_1 =
        3\), \(v_2 = 4\), and \(v_3 = 2\).
    \item Once the shape of \(M\) is defined and all of the elements have been
        populated, the shape will not change and all of the elements will remain
        populated. For example, the aforementioned \(3 \times 4 \times 2\) array
        must always contain 3 arrays, each of which contains 4 arrays of 2
        elements.
    \item Each \(i\)-the dimension from \(1\) to \(d-1\) must contain \(v_i\)
        elements, each of which is an array of length \(v_{i+1}\); in other
        words, the length of each element in a particular dimension must be
        identical to one another.
\end{itemize}

In summary, these properties describe the notion of a ``well-formed'' matrix
that ensures that the size \(v_i\) of each dimension remains fixed and that all
elements in a particular dimension \(i\) have the same length as one another.
This property is generally true of matrices, hence the references to both an
``arrays'' and ``matrices'' in this section. Unless otherwise noted, ``array''
refers to the generic \matlab\ object and ``matrix'' refers to an ``array'' with
these properties when discussed in the scope of this paper.

\subsection{Operations}

\matlab\ supports a number of operations that can be used to manipulate arrays,
but only a subset are used in our program. To illustrate the operations used, we
begin with a visualization of a three-dimensional array \(E\) with dimensions
\(3 \times 4 \times 2\) below and discuss how it is used in \matlab.

@@

\begin{center}
    \begin{tikzpicture}[every node/.style={anchor=north east,fill=white, minimum width=.7cm}]
        \matrix (mA) [draw,matrix of math nodes]
        {
            1 & 3 & 5 & 7 \\
            2 & 4 & 6 & 8 \\
        };

@@

        \draw[dashed](mA.north east)--(mC.north east);
        \draw[dashed](mA.north west)--(mC.north west);
        \draw[dashed](mA.south east)--(mC.south east);
    \end{tikzpicture}
\end{center}

\subsubsection{Element Access}

For the matrix \(E\), each element would be indexed according to the following
mapping. Under this scheme, for example, the element \(8\) in \(E\) would
have the index \((1, 4, 2)\). This indexing scheme is the general pattern used to
access one particular element out of a multi-dimensional array. Note that arrays
in \matlab\ have indices starting at 1. The indexing scheme is demonstrated
below.

\begin{center}
    \begin{tikzpicture}
        \def\xs{1} %shift in x direction
        \def\ys{1.5} %shift in y direction
        \def\nm{3} % number of 2d matrices in the 3d matrix
        \foreach \x in {1,2,...,\nm}
        {
            \matrix [draw, % for the rectangle border
            fill=white, % so that it is not transparent
            ampersand replacement=\&] %see explanation
            (mm\x)%give the matrix a name
            at(-\x * \xs, -\x * \ys) %shift the matrix
            {
                \node {(\x,1,1)}; \& \node {(\x,2,1)}; \& \node {(\x,3,1)}; \& \node {(\x,4,1)};\\
                \node {(\x,1,2)}; \& \node {(\x,2,2)}; \& \node {(\x,3,2)}; \& \node {(\x,4,2)};\\
            };
        }

        \draw [dashed,gray](mm1.north west) -- (mm\nm.north west);
        \draw [dashed,gray](mm1.north east) -- (mm\nm.north east);
        \draw [dashed,gray](mm1.south east) -- (mm\nm.south east);
    \end{tikzpicture}
\end{center}

\subsubsection{Linearized Indexing}\label{S:linearized_indexing}

Note that in \(E\), elements are ordered in ascending order with respect to
their indexing. First, sub-array \((1)\) is populated, and \((1, 1)\) is the
first array within it, so it is filled with elements \(1\) and \(2\). Array
\((1, 2)\) is then populated by elements \(3\) and \(4\). After all four of
\(E\)'s sub-arrays have been populated, sub-array \((2)\) is populated, and so
on until the array is full.

Given the numbering scheme in \(E\), one can imagine a ``flattened'' version of
\(E\) that is an one-dimensional array containing all of \(E\)'s elements;
simply put, an array containing the elements \(1\) to \(24\) in numerical
order. \matlab\ actually allows arrays of arbitrarily many dimensions to be
logically flattened during accesses so that elements can be accessed with one
index, like in a one-dimensional array. We found this property to be somewhat
interesting and challenging to model, and we refer to the flattening of a matrix
as ``linearization.''

The process of linearizing a matrix is perhaps better explained with a
two-dimensional example. Imagine a general two-dimensional \(3 \times 4\) matrix,
such as the one below, where:
\begin{equation*}
    \begin{bmatrix}
        a & b & c & d \\
        e & f & g & h \\
        i & j & k & l
    \end{bmatrix}
\end{equation*}
To linearize the matrix, we label each value with an increasing index from the
top-left, left-to-right, top-to-bottom. For example:
\begin{equation*}
    \begin{bmatrix}
        a_1 & b_2 & c_3 & d_4 \\
        e_5 & f_6 & g_7 & h_8 \\
        i_9 & j_{10} & k_{11} & l_{12}
    \end{bmatrix}
\end{equation*}
Finally, we create a vector, a one-dimensional matrix, by reading the elements
off in order:
\setcounter{MaxMatrixCols}{20}
\begin{equation*}
    \begin{bmatrix}
        a & b & c & d & e & f & g & h & i & j & k & l
    \end{bmatrix}
\end{equation*}
This process generalizes to higher dimensions. The length of the resulting
vector is always the product of the sizes of the dimensions. To see this, note
that this product counts the number of elements in the matrix and that the
linearized matrix has the same number of elements in a single dimension.

Under this scheme, any element in a matrix can be accessed via its position in
the linearized version of that matrix. For example, the element \(12\) in \(E\)
can be accessed by the index \((12)\) and in the previous example, the element
\(f\) can be accessed by the index \((6)\).

\subsubsection{Ranged Indexing}\label{S:ranged_indexing}

Broadly speaking, \matlab\ supports the notion of a range of numbers specified
by a starting index and a stopping index, which includes every index between the
starting index and the stopping index (inclusive). This abstraction can be used
to access or assign elements within a sub-matrix of a particular matrix. For
example, suppose in matrix \(E\) you only cared about the first matrix in the
first dimension; the second, third, and fourth columns within the second
dimension; and the second row within the third dimension. This sub-matrix can be
specified by the index \((1, 2:4, 2)\) and would be the matrix

\begin{equation*}
    \begin{bmatrix}
        4 & 6 & 8
    \end{bmatrix}.
\end{equation*}

Suppose that you wanted to get the second, third, and fourth columns within the
second dimension and the second row within the third dimension for \emph{all
three} of the outer-most dimension of \(E\) but did not know \textit{a priori}
how many elements it had. \matlab\ supports an ``inclusive range'' operator,
represented by a colon \texttt{:}. This operator represents the range
encompassing all values in a particular dimension. The sub-matrix described here
would be specified by the index $(:, 2:4, 2)$ and represents the following
matrix:

\begin{center}
    \begin{tikzpicture}[every node/.style={anchor=north east,fill=white, minimum width=.7cm}]
        \matrix (mA) [draw,matrix of math nodes]
        {
            4 & 6 & 8 \\
        };

@@

        \draw[dashed](mA.north east)--(mC.north east);
        \draw[dashed](mA.north west)--(mC.north west);
        \draw[dashed](mA.south east)--(mC.south east);
    \end{tikzpicture}
\end{center}

\subsubsection{Matrix Operations}\label{S:matrix_operations}

Details about \matlab\ arrays pertaining to indexing are perhaps some of their
most interesting features, but we also use some typical matrix operators in our
program. Generally speaking, matrix operators can be generalized to any number
of dimensions, as long as the dimensions of the arrays agree for the operation
in question. There are two broad classes of matrix operators that we make use
of:

\begin{description}
    \item[Element-wise] For two matrices of the same dimensionality, the
        operator is applied on each of the two elements containing the same
        index in the two matrices.
    \item[All Elements] For an array and a scalar, the operation and scalar
        (\textit{e.g.}, add 1) is applied to every element in the matrix.
\end{description}

\subsection{Modeling the Problem}

Aside from the core matrix operations, we anticipated that we would need to
model the state of our program at all points of execution since our goal was to
prove termination. To this end, we would need to identify all of the syntax that
we use in our program and specify their associated semantics. The upshot of
modeling our program in this way would be the ability to remains to prove its
termination of our given that certain preconditions hold.

@@

When we started this project, we had in mind that the preconditions would be
statements about the program's inputs; for example, ``the length and width of
key art are always multiples of the thumbnails' size.'' Ultimately, we did not
get far enough in modeling our program to devise a set of pre-conditions, but
the point remains that we needed to be able to model the control flow of our
entire program in order to do this.

\section{Design and Approach}\label{S:design}

@@

Much of the abstract syntax is relatively self-explanatory, so complete
definitions are elided here. However, we wish to draw attention to some of the
less apparent and more interesting syntax present in \mmatlab:

@@

    \item \textsf{AddExp}: While this is nominally a regular addition operator,
        \matlab's addition is interesting in that it has overloaded parameter
        types and expected behavior for each. The addition of two numbers is a
        number and the addition of two matrices results in element-wise
        addition, as one might expect. However, the sum of a number and a
        matrix, is a matrix with all elements incremented by the number. Similar
        behaviors hold for \textsf{MultExp} and \textsf{DivExp}.

    \item \textsf{RangeExp}: Like Python, \matlab\ has an expression where a
        starting number and stopping number may be provided, generating an
        inclusive list of all numbers between them. This is discussed in
        \S~\ref{S:ranged_indexing}.

    \item \textsf{IndexExp}: This expression pairs a variable with an indexer,
        which can be an almost arbitrary expression. \matlab\ does restrict the
        syntactic placement of certain indexing operations.

@@

As mentioned in \S~\ref{S:design}, our original formulation of matrices as
nested dependent \texttt{Vector.t}s quickly grew in complexity. We abandoned
that approach in favor of a type that is not guaranteed to behave well, but
whose definition makes defining operations more natural~\cite{SO_2021_2}. This
approach removes the proofs that dependent types provide automatically, so we
have to give those ourselves.

@@

\paragraph{Linearizing a matrix} See \S~\ref{S:linearized_indexing} for more
details on matrix linearization. To define linearization in Coq, we use a
two-stage unpacking as with other matrix operations. The function
\(\func{linearize'}\) turns scalars into a singleton list and recursively
flat-maps itself over matrices to produce the one-dimensional matrix as a list.
The function \(\func{linearize}\) unpacks a matrix, then transforms the contents
by wrapping the result of mapping \(\func{linearize'}\) into \(\func{Scalar}\)s
in a \(\func{Matrix}\) and transforms the shape by taking the product of the
(normalized) shape of the matrix. Recall that the length of the linearized
vector is the product of the sizes of all the dimensions. The definitions are
spelled out below:

@@

\paragraph{Linearly indexing a matrix} Given the operation to linearise a
matrix, one-based linearly indexing is easy. See \S~\ref{S:linearized_indexing}
for details on linear indexing. The function \(\func{nth}\) takes a matrix \(m\)
and an index \(n\); it returns the one-based linearly indexed content of \(m\)
at \(n\) (wrapped in the option constructor \(\func{Some}\)) if \(n\) is in
bounds relative to \(m\). When \(n\) is out of bounds, the result is
\(\func{None}\). For one-based indexing with functions that naturally use
zero-based indexing, we use \(n-1\). We should note that the current version of
\(\func{nth}\) has the subtle bug that indexing at \(n=0\) currently returns the
same as indexing at \(n=1\), when it should in fact be out of bounds and thus
\(\func{None}\)---a simple \texttt{if} or \texttt{match} guard would cover this
case. We note that our linear indexing supports only a single index, while
\matlab\ allows any index prior to the last to start linearization.

@@
